Rheology of Cubic Blue Phases

### 2012-10-11 papers/bprheo/bprheo.tex
New draft BP rheology paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1790 40876441-3496-0410-a613-a2fca35b2ae2

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a cholesteric helix is subjected to a Poiseuille flow along
the helical axis, small pressure differences drive flow mainly through
``permeation'', first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils and 
leaves a flow-induced nematic phase.
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

Dupuis et al.~\cite{Dupuis:2005} performed the first numerical investigation of BP rheology 
in Poiseuille flow, starting from equilibrium structures of BPI and BPII and a periodic 
array of doubly twisted cylinders.
Under small forcing, the network opposed the flow giving rise to a significant 
increase in apparent viscosity.
Upon increasing the forcing they found clear evidence of shear thinning.
In the crossover region they predicted a novel oscillatory regime where the network 
continuously breaks and reforms as portions of the disclinations in the center of the channel 
move to neighboring cells and relink with the parts of the network left behind by the flow. 
Compared with the cholesteric case, the viscosity still decreases with forcing 
(the system shear thins) but much less than
for cholesterics in the permeation mode, which is in agreement with experiments
~\cite{Zapotocky:1999, Ramos:2002}.

@@

Our approach is based on the well-established Beris-Edwards model for hydrodynamics of
cholesteric liquid crystals \cite{Beris:1994}, which describes the ordered state 
in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
In the uniaxial approximation, the order parameter is given by
$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
order. More generally,
the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
characterizes the local degree of orientational order.
The thermodynamic properties of the liquid crystal are determined by a free energy
${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2\nonumber\\
&-&\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
\end{eqnarray}
The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to~\cite{Alexander:2006}.

@@

The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation accoring to
\begin{equation}
X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N),
\label{spectrum}
\end{equation}
wheras we subtracted the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
$N$ was the number of points in the time series, typically one for every 100 LB time steps.

@@

\begin{table*}[htpb]
\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
\hline
$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
\hline
BPI \\
\hline
0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
\hline
BPII \\
\hline
0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
\hline
\end{tabular}
\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding any transient dynamics.}
\label{tab1}
\end{table*}

@@

As usual in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our runs with 
analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit 
and equilibrated these configuration for 5000 LB timesteps before we started the shear flow. 
During the equilibration sequence the optimal redshift $r$ was calculated and applied at every timestep.
After equilibration the redshift was kept constant throughout the rest of the simulation with shear flow.
We chose a pitch length of 32 LBU for BPII and 64 LBU in case of BPI, which is the equivalent 
of 4 unit cells along each coordinate direction and a total of 64 unit cells in our simulation box.
Runs with higher resolution confirmed that the resolution was adequate to track  
all kinematic details of the blue phase networks in shear flow.

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction and the 
velocity gradient along the y-direction.
We applied a correction for spurious momentum currents that emerge due to required interpolations across 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
For clarity we classify various subregimes, two in case of BPII and three in case of BPI. 
They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (top) and strain (bottom). 
For clarity the individual curves have been shifted along the time and strain axis. 
The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
regular periodic breakup and reconnecting of the discination lines (BPII-1, open circles) 
and break up of the network into a travelling helical wave (BPII-2, solid circle).}
\label{bp2-rheo}
\end{figure}

We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.
Ahead of the more detailed discussion we provide first a general overview of the flow behaviour at all applied shear rates.
Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
deviatoric stress normalised to the isotropic background viscosity: 
$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
any additional contribution from the liquid crystal.
The top picutre shows the date over time, whereas the graph at the bottom gives 
the same data versus accummulated strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.

@@

Interestingly, while being displaced with the flow the entire network moves perpendicular 
to the flow and velocity gradient in the vorticity direction.
A similar behaviour has been recently observed for blue phases in shear flow and 
confined geometries ~\cite{Henrich:2012b}.
But contrary to the stick-slip motion that has been reported there the movement is 
steady. It occurs in such a manner that the positions of breakup and reconnection 
of the network, visible in Fig. \ref{bp2-med}, are slightly offset and make the network 
travel along the z-direction.
The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
i.e. it takes a displacement of six unit cells in flow directions 
until the network has moved one unit cell in voriticity direction.

@@

Characteristic bands are visible, which are oriented along the vorticity direction and
normally hidden behind the primary flow component.
When the helicity of the underlying cholesteric phase is changed from left- to a 
righthanded the pattern becomes a mirror image of the former one. The sign of 
$v_z$ change and also the sense of motion is inverted.
Further quantitative evidence for a direct link between the sense of motion 
and the helicity can be gained by time-averaging over individual cycles.
Tab. \ref{tab1} gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the z-components.
There is only a slight imbalance between the maximum and minimum velocities 
in z-direction, suggesting that the movement of the network in vorticity 
direction is permeative. Note that this imbalance is retained under inversion of
helicity.

@@

If the flow rate exceeds a certain critical value BPII does not undergo the 
affine transformation descibed above and adopts another mode of flow. 
Fig. \ref{bp2-high} shows a sequence of cuts through the director 
field at high shear rate. The entire blue phase network has broken up 
into a simple cholesteric helix and flows as a travelling 
helical wave with the helical axis oriented along the vorticity direction.
The state is translationally invariant in flow and gradient direction.
and is the preferred mode of flow of a cholesteric 
liquid crystal at low Ericksen numbers below the uncoiling transition 
to a nematic, flow-aligned state at even higher shear rates ~\cite{Rey:1996a, Rey:1996b}.

In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$.
At the moment we do not have a hard criterium for the critical
shear rate at hand. However, it seems reasonable that the critical shear rate 
depends on thermodynamic and kinetic parameters such as the effective 
temperature $\tau$, the chirality $\kappa$ and the rotational diffusion 
constant $\Gamma$.

@@

In regime BPI-2 at lower shear rates periodic oscillations emerge in the 
deviatoric stress that are not sinusoidal, but show an increasingly complex
pattern up to the breakup point.
For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
they become  unregular and finally disappear after a few
cycles. Shear-thinning is slightly stronger than in BPII. 
The average shear stress during one cycle can be fitted reasonably 
well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.

@@

The rheological response of BPI at low shear rate $\gd<3.91\e{-5}$ 
is strikingly different from that of BPII and exhibits a
dynamic yield stress.
An explanation for this strange behaviour can be found by 
looking more closely at the average deviatoric stress and 
free energy density, as shown in Fig. \ref{bp1-fe-yield}.
When the quiescent and equilibrated BPI network begins to flow
the shear stress increases steeply and goes through a global maximum.
Shortly after it goes negative (the total shear stress is still
positive as the backgound viscosity has been subracted) and positive
again, forming a complete cycle. 
This cycle is repeated for a couple of times before 
the amorphous configuration emerges and any regular pattern
disappears.

@@

The negative branch in the deviatoric stress is connected
to a local maximum followed by a local minimum in the average free energy.
Hence, the region of negative stress can be 
regarded as a phase during which elastic forces excert a 
'pull' onto the network while it tries to reach a more 
favourable configuration. However, the low flow rate
causes this unstable intermediate state to be exposed to these
pulling elastic forces for a relatively long time, long enough to 
create small distortions which finally break the symmetry of the
flow-induced configuration.

@@

Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) which features yield stress 
behaviour at very low flow rates and the onset of power-law behaviour 
at slightly larger shear rates (viz. Fig. \ref{bp1-rheo}) lies another 
region ($3.91\e{-5}\le\gd< 6.25\e{-4}$) where the network flows with periodically 
recurring patterns. The transition between these two regimes is shown
in Fig. \ref{bp1-1_bp1-2} in terms of deviatoric stress over total strain.

The reason for this qualitative difference is rather obscur. A possible 
explanation is that due to the larger flow rate the network kinetically 
bypasses the relatively large stress fluctuations during one cycle which
are typical in regime BPI-1. For even lower $\gd$ the deviatoric stress
goes actually negative (viz. Fig. \ref{bp1-fe-yield}). This destabilises
the periodic network very quickly and leads to an amorphous configuration
in the steady state.

Fig. \ref{bp1-med} shows snapshots of BPI in steady shear flow. 
Contrary to BPII the flow-induced conformations of BPI differ significanly 
from the appearance of BPI at equilibrium. The depicted 
state is rather surprisingly still kinematically connected to the equilibrium state.
When the shear flow is switched off the flow-induced configuration reverts
to its original configuration of a BPI network at rest.
%%%%%%%%%%%%
%The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.
%%%%%%%%%%%
After each cycle the position of the network is displaced with respect to the beginning of the cycle.
However, neither the magnitude nor the direction of the offset obey a clear rule,
are different and non-integer for every shear rate in this regime of flow. 
This suggests the described phenomenon constitutes an instance of rheochaos in this 
regime of liquid crystalline flow.

>>>>

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a cholesteric helix is subjected to a Poiseuille flow along
the helical axis, small pressure differences drive flow mainly through
''permeation'', first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils and 
leaves a flow-induced nematic phase.
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

Dupuis et al.~\cite{Dupuis:2005} performed the first numerical investigation of BP rheology 
in Poiseuille flow, starting from equilibrium structures of BPI and BPII and a periodic 
array of doubly twisted cylinders.
Under small forcing, the network opposed the flow giving rise to a significant 
increase in apparent viscosity.
Upon increasing the forcing they found clear evidence of shear thinning.
In the crossover region they predicted a novel oscillatory regime where the network 
continuously breaks and reforms as portions of the disclinations in the centre of the channel 
move to neighbouring cells and relink with the parts of the network left behind by the flow. 
Compared with the cholesteric case, the viscosity still decreases with forcing 
(the system shear thins) but much less than
for cholesterics in the permeation mode, which is in agreement with experiments
~\cite{Zapotocky:1999, Ramos:2002}.

@@

Our approach is based on the well-established Beris-Edwards model for hydrodynamics of
cholesteric liquid crystals \cite{Beris:1994}, which describes the ordered state 
in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
In the uniaxial approximation, the order parameter is given by
$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
order. More generally,
the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
characterises the local degree of orientational order.
The thermodynamic properties of the liquid crystal are determined by a free energy
${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2\nonumber\\
&-&\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
\end{eqnarray}
The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to~\cite{Alexander:2006}.

@@

The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
\begin{equation}
X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N),
\label{spectrum}
\end{equation}
whereas we subtracted the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
$N$ was the number of points in the time series, typically one for every 100 LB time steps.

@@

\begin{table*}[htpb]
\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
\hline
$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
\hline
BPI \\
\hline
0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
\hline
BPII \\
\hline
0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
\hline
\end{tabular}
\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minimum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding any transient dynamics.}
\label{tab1}
\end{table*}

@@

As usual in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our runs with 
analytical solutions that minimise the free energy functional Eq.\ref{FE} in the high-chirality limit 
and equilibrated these configuration for 5000 LB timesteps before we started the shear flow. 
During the equilibration sequence the optimal redshift $r$ was calculated and applied at every timestep.
After equilibration the redshift was kept constant throughout the rest of the simulation with shear flow.
We chose a pitch length of 32 LBU for BPII and 64 LBU in case of BPI, which is the equivalent 
of 4 unit cells along each coordinate direction and a total of 64 unit cells in our simulation box.
Runs with higher resolution confirmed that the resolution was adequate to track  
all kinematic details of the blue phase networks in shear flow.

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction and the 
velocity gradient along the y-direction.
We applied a correction for spurious momentum currents that emerge due to required interpolations across 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
For clarity we classify various sub-regimes, two in case of BPII and three in case of BPI. 
They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (top) and strain (bottom). 
For clarity the individual curves have been shifted along the time and strain axis. 
The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
and break up of the network into a travelling helical wave (BPII-2, solid circle).}
\label{bp2-rheo}
\end{figure}

We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.
Ahead of the more detailed discussion we provide first a general overview of the flow behaviour at all applied shear rates.
Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
deviatoric stress normalised to the isotropic background viscosity: 
$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
any additional contribution from the liquid crystal.
The top picture shows the date over time, whereas the graph at the bottom gives 
the same data versus accumulated strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.

@@

Interestingly, while being displaced with the flow the entire network moves perpendicular 
to the flow and velocity gradient in the vorticity direction.
A similar behaviour has been recently observed for blue phases in shear flow and 
confined geometries ~\cite{Henrich:2012b}.
But contrary to the stick-slip motion that has been reported there the movement is 
steady. It occurs in such a manner that the positions of breakup and reconnection 
of the network, visible in Fig. \ref{bp2-med}, are slightly offset and make the network 
travel along the z-direction.
The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
i.e. it takes a displacement of six unit cells in flow directions 
until the network has moved one unit cell in vorticity direction.

@@

Characteristic bands are visible, which are oriented along the vorticity direction and
normally hidden behind the primary flow component.
When the helicity of the underlying cholesteric phase is changed from left- to a 
right-handed the pattern becomes a mirror image of the former one. The sign of 
$v_z$ change and also the sense of motion is inverted.
Further quantitative evidence for a direct link between the sense of motion 
and the helicity can be gained by time-averaging over individual cycles.
Tab. \ref{tab1} gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the z-components.
There is only a slight imbalance between the maximum and minimum velocities 
in z-direction, suggesting that the movement of the network in vorticity 
direction is permeative. Note that this imbalance is retained under inversion of
helicity.

@@

If the flow rate exceeds a certain critical value BPII does not undergo the 
affine transformation  above and adopts another mode of flow. 
Fig. \ref{bp2-high} shows a sequence of cuts through the director 
field at high shear rate. The entire blue phase network has broken up 
into a simple cholesteric helix and flows as a travelling 
helical wave with the helical axis oriented along the vorticity direction.
The state is translationally invariant in flow and gradient direction.
and is the preferred mode of flow of a cholesteric 
liquid crystal at low Ericksen numbers below the uncoiling transition 
to a nematic, flow-aligned state at even higher shear rates ~\cite{Rey:1996a, Rey:1996b}.

In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$, which
corresponds to Ericksen numbers of ${\cal E}\simeq10$, based on our
parameters ($K=0.02, \eta=0.66$), typical velocities (viz. Tab. \ref{tab1}) 
and length scales ($p/2$).

@@

In regime BPI-2 at lower shear rates periodic oscillations emerge in the 
deviatoric stress that are not sinusoidal, but show an increasingly complex
pattern up to the breakup point.
For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
they become  irregularunregular and finally disappear after a few
cycles. Shear-thinning is slightly stronger than in BPII. 
The average shear stress during one cycle can be fitted reasonably 
well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.

@@

The rheological response of BPI at low shear rate $\gd<3.91\e{-5}$ 
is strikingly different from that of BPII and exhibits a
dynamic yield stress.
An explanation for this strange behaviour can be found by 
looking more closely at the average deviatoric stress and 
free energy density, as shown in Fig. \ref{bp1-fe-yield}.
When the quiescent and equilibrated BPI network begins to flow
the shear stress increases steeply and goes through a global maximum.
Shortly after it goes negative (the total shear stress is still
positive as the background viscosity has been subtracted) and positive
again, forming a complete cycle. 
This cycle is repeated for a couple of times before 
the amorphous configuration emerges and any regular pattern
disappears.

@@

The negative branch in the deviatoric stress is connected
to a local maximum followed by a local minimum in the average free energy.
Hence, the region of negative stress can be 
regarded as a phase during which elastic forces exert a 
'pull' onto the network while it tries to reach a more 
favourable configuration. However, the low flow rate
causes this unstable intermediate state to be exposed to these
pulling elastic forces for a relatively long time, long enough to 
create small distortions which finally break the symmetry of the
flow-induced configuration.

@@

Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) at 
at slightly larger shear rates $3.91\e{-5}\le\gd< 6.25\e{-4}$
and Ericksen numbers $0.5\le{\cal E}<10$ 
lies another region where the network flows with periodically 
recurring flow-induced conformations (viz. Fig. \ref{bp1-rheo}). 
We refer to this region as BPI-2. The transition between BPI-1 and BPI-2 
is shown in Fig. \ref{bp1-1_bp1-2} by means of the deviatoric stress over total strain. 
Obviously a qualitative difference between these two is the absence of 
large stress fluctuations that occur during one typical cycle in regime BPI-1.
Remember that for even lower $\gd$ the deviatoric stress becomes 
temporarily negative (viz. Fig. \ref{bp1-fe-yield}), which 
caused destabilisation of the periodic network and led to the 
amorphous configuration in the steady state. Hence, an explanation 
for this phenomenon could be that in BPI-2 the network finds a 
kinetic path to bypass these large amplitude fluctuations and 
prevents the network from destabilising.

Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
BPI in steady shear flow. Contrary to BPII-1 at these shear rates
, BPI-2 does not resemble the equilibrium configuration
undergoing an affine transformation. It features rather intricate, flow-induced 
conformations. Surprisingly this state is kinetically connected to the equilibrium 
state as when the shear flow is switched off the flow-induced configuration 
reverts to a perfect quiescent BPI unless it got stuck in a local free energy 
minimum.
